### 2013-01-05 edupar-13/410.tex
410.tex: Added Pebbles kspec summary.

\subsubsection{x86}
Students write code that runs on one specific
hardware platform.
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them boot
up their code, type commands, and observe the results.
Also, unlike many System-on-a-Chip platforms,
detailed hardware descriptions are readily available.
% rather than being proprietary or "documented"
% by labyrinthine Linux source code
Finally, the platform is fairly stable---with only a
little care, it is possible to write a compliant
kernel which runs on anything newer than a 486DX cpu,
i.e., almost any PC hardware which is likely to be in
working condition.

\subsubsection{32-Bit} The course still targets the 32-bit
subset of the x86 platform, for three reasons.
%% TODO I'm not sure I believe this is an advantage. Instead of having
%% to verify if each register is correct, students now have to verify
%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
First, while the small number of registers is
frustrating for compiler writers,
from the point of view of OS students,
the existence of fewer registers means less work
when it is necessary to verify that each register's
value is correct.
Likewise, the old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater insight.
Finally, since it is impossible for an x86 processor
to be in 64-bit mode with virtual memory off,
students would need to be coached through the
32-bit--to--64-bit transition or would need
to be given a template virtual-memory implementation.
All in all, at present we are satisifed with
the 32-bit choice.

\subsubsection{Memory Model} At present we allow students
to reason about concurrency based on a
sequentially-consistent view of
memory, which is consistent with
existing x86 hardware~\cite{SewellSOZNM:x86tso-cacm10}
but simplistic compared to most other
architectures.
Lecture material educates students about the
general topic of out-of-order memory systems
and fencing,
but at present the hands-on experience does not
reinforce that material.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Pebbles}
\label{sec:pebbles}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% primeval Unix model: fork()/exec()/wait()
% Mach ``task/thread'' terminology/structure (as opposed to rfork()/clone())
% keyboard input, screen output
% thread creation
% thread block/unblock
% new_pages()/remove_pages()
% swexn

% swexn - what an auto-growing stack really means

>>>>

\subsubsection{x86}

Students write code that runs on one specific
hardware platform.
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them boot
up their code, type commands, and observe the results.
Also, unlike many System-on-a-Chip platforms,
detailed hardware descriptions are readily available.
% rather than being proprietary or "documented"
% by labyrinthine Linux source code
Finally, the platform is fairly stable---with only a
little care, it is possible to write a compliant
kernel which runs on anything newer than a 486DX cpu,
i.e., almost any PC hardware which is likely to be in
working condition.

\subsubsection{32-Bit}

The course still targets the 32-bit
subset of the x86 platform, for three reasons.
%% TODO I'm not sure I believe this is an advantage. Instead of having
%% to verify if each register is correct, students now have to verify
%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
First, while the small number of registers is
frustrating for compiler writers,
from the point of view of OS students,
the existence of fewer registers means less work
when it is necessary to verify that each register's
value is correct.
Likewise, the old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater insight.
Finally, since it is impossible for an x86 processor
to be in 64-bit mode with virtual memory off,
students would need to be coached through the
32-bit--to--64-bit transition or would need
to be given a template virtual-memory implementation.
All in all, at present we are satisifed with
the 32-bit choice.

\subsubsection{Memory Model}

At present we allow students
to reason about concurrency based on a
sequentially-consistent view of
memory, which is consistent with
existing x86 hardware~\cite{SewellSOZNM:x86tso-cacm10}
but simplistic compared to most other
architectures.
Lecture material educates students about the
general topic of out-of-order memory systems
and fencing,
but at present the hands-on experience does not
reinforce that material.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Pebbles kernel specification}
\label{sec:pebbles}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The kernel that students implement is loosely based on the
Unix model: \texttt{fork()}, \texttt{exec()}, and \texttt{wait()}.
We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
distinction between ``tasks,''
which are resource containers,
and threads,
each of which executes within a single task.
This partitioning is a different approach from that of
Plan~9's \texttt{rfork()} or Linux's \texttt{clone()},
which allow a continuum of sharing between
identical threads and independent processes,
at the cost of implementation complexity which we wish to avoid.
A task's address space can be edited through two system
calls that add and delete contiguous regions of memory.
Students are expected to implement zero-fill-on-demand,
but copy-on-write and demand-loading of executables are
optional.
Threads can print text to and read lines of text from
the simple console that students implemented at
the start of the semester.

Within a task, threads are created using the
\texttt{thread\_fork} system call, which closely
mirrors the specification of \texttt{fork()}:
the new thread begins with an exact copy of the
register values of the old thread, except that
the return-value register, \texttt{\%eax}, contains
zero;
in the old thread, the return value is the thread i.d.\
of the newly created thread.
This minimal formulation defers all policy decisions
(e.g., stack size, location, and contents) to the
user-space thread library.
A thread may suspend its execution via the
\texttt{deschedule()} system call,
and a blocked thread can be reactivated via
\texttt{make\_runnable()}.

Each thread has the ability to register a handler
for hardware exceptions (illegal instruction,
division by zero, etc.)
via the \texttt{swexn()} system call,
which specifies the addresses of handler code
and a handler stack.
When a thread encounters an exception,
the kernel reflects the exception to user space
by
de-registering the handler,
pushing the thread's register state onto the stack,
and invoking the handler code.
The handler has the ability to re-register itself
(or register some other handler),
to run arbitrary code,
and/or to atomically adopt a new set of register
values.
The \texttt{swexn()} facility is used to enable
the thread library to detect when a managed
thread has crashed
and, more importantly,
to enable user-space
implementation of automatically-growing stack
regions.
Removing the auto-stack policy and mechanism from
its traditional position in Unix-like kernels
was done because it simplfies the kernel and
``front-loads'' the policy question to an earlier
part of the course,
before students are facing the substantial pressure
of the kernel project.
